---
title: "A visual introduction to rollply"
author: "Alexandre Génin <alex@lecairn.org>"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{rollply}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# A visual introduction to rollply


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r, output = "none", messages = FALSE}
library(proj4) 
library(ggplot2) 
library(rgdal)
library(tidyr) 
library(rollply) 
library(plyr) 
```

Rollply is a small function built upon [plyr](http://plyr.had.co.nz/)'s `ddply` 
function to facilitate moving-window-based computations. If you have a 
`data.frame`, give the dimensions over which the window should 
move, and rollply will make the subsets, apply the function on them and then 
combine the results into a data.frame.

In short, rollply extends the 
[split-apply-combine strategy](http://plyr.had.co.nz/) to moving-window 
computations, using a similar syntax. This tutorial thus assumes some basic 
familiarity with `plyr`, and in particular the function `ddply`. 
Let's start with a simple example.

# Examples

## A time-series example

A simple use of moving-windows is adding a trendline to a time series plot.
We will use the CO2 data from the Mauna Loa [NOAA
Observatory](http://www.esrl.noaa.gov/gmd/ccgg/trends/) as a
environmentally-conscious example.

```{r, fig.width=7, fig.height=4}
# Download and format data
# FTP access does not work on Travis so we use a copy of this dataset
# url <- "ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt"
url <- "https://alex.lecairn.org/co2_mm_mlo.txt"
hawaii <- read.table(url)[ ,c(3,4)]
names(hawaii) <- c('date','CO2')
hawaii[hawaii$CO2 < 0, "CO2"] <- NA # mark NAs as such

# Display original trend
CO2.plot <- ggplot(hawaii) + geom_line(aes(date, CO2)) + ylab("CO2 (ppm)")
print(CO2.plot)
```

There is a clear trend here! Let's smooth out the season effect (the wiggles
in the black curve). We'll use a window with a size of one year to compute a
yearly average.

```{r, fig.width=7, fig.height=4}
# with smoothed trend
hawaii.smoothed <- rollply(hawaii, ~ date, wdw.size = 1,
                           summarize, CO2.mean = mean(CO2, na.rm = TRUE), )
CO2.plot + geom_line(aes(date, CO2.mean), data = hawaii.smoothed, color = 'red')
```

And voilà! A rather nice, although a bit depressing trend line for our data. 
When working on time-series, this represents an alternative to specialized 
packages such as `zoo` that also provide tools to apply functions over rolling
windows. 

Let's take a more complex example that works on two-dimensional data.

## Exploring french town names

If you open a map of France, you'll notice that towns and villages
tend to have names that follow patterns. For example, Brittany's towns are
famous for having names starting with a *"ker-"*. Many towns in
[Lorraine](https://en.wikipedia.org/wiki/Lorraine_%28region%29) end in *"-ange"*
(a legacy from the german ending *"-ingen"*).

Can we visually explore the distribution of french towns names ? 
rollply can help here.

A moving-window approach essentially boils down to the following steps:

1. Build a grid over the whole country
2. For each point of the grid, take all the data poitns (towns) that are less 
than xx kilometers from it.
3. Check the names of the towns and count the ones matching a pattern. Return
the number of matching towns.
4. Combine the results in a `data.frame`

Like `ddply` (in package `plyr`), `rollply` takes care of points 1,2 and 4. We
just need to define a function that does number 3. 
 
Let's download a [dataset](http://www.nosdonnees.fr/dataset/listes-des-communes-par-r-gions-d-partements-circonscriptions) of town names with their geographical coordinates:

```{r, warning=FALSE, results="hide"}
# Download and prepare dataset
# Source and decription: 
# https://publicdata.eu/dataset/listes-des-communes-par-rgions-dpartements-circonscriptions

tmpfile <- tempfile()
url <- paste0('http://www.nosdonnees.fr/wiki/images/b/b5/',
         'EUCircos_Regions_departements_circonscriptions_communes_gps.csv.gz')
download.file(url, destfile = tmpfile)
dat <- read.csv2(tmpfile, stringsAsFactors = FALSE)
file.remove(tmpfile)
dat <- dat[with(dat, latitude != "" | ! grepl(",", latitude)), 
           c('nom_commune', 'latitude', 'longitude')]
colnames(dat) <- c('name', 'latitude', 'longitude')

dat[ ,'name']      <- as.factor(tolower(dat[ ,'name']))
dat[ ,'latitude']  <- as.numeric(dat[ ,'latitude'])
dat[ ,'longitude'] <- as.numeric(dat[ ,'longitude'])

# We use an equirectangular projection to work on true distances
dat <- na.omit(dat)
dat <- data.frame(dat, proj4::project(dat[ ,c('longitude','latitude')],
                                      '+proj=eqc'))
dat <- dat[ ,c('name','x','y')]

```

```{r, fig.width=7, fig.height=6}
# Visualise distribution of towns
str(dat)
ggplot(dat) + geom_point(aes(x, y), alpha=.1)
```

Nice, let's see whether *"ker"*-named towns mainly occur in Brittany. 

```{r, fig.width=7, fig.height=6}

# This is our custom function : it accepts a data frame and a regular 
# expression and returns the number of matches in the column "name", formatted
# withing a data.frame (this plays well with ddply).
how_many_with_name <- function(df, regexp) { 
  data.frame(ker = sum(grepl(regexp, df[ ,'name'])))
}

dat_roll <- rollply(dat, ~ x + y, wdw.size = 1e4, grid_npts = 10e3,
                    how_many_with_name, regexp = "^ker")

ggplot(dat_roll) +
  geom_raster(aes(x, y, fill = ker)) +
  scale_fill_distiller(palette = 'Greys')
```

It seems there are indeed many towns with a name starting with "*ker*" in
Brittany (and a couple in Alsace/Lorraine, too). However, our plot is pretty
ugly: we cannot see the actual country shape! When nothing is specified,
rollply computes its values over a rectangular grid than spans the maximum width
and height of the original dataset.

Here, the spatial distribution of towns reflects pretty nicely the overall shape 
of the country, so instead of a building a rectangular grid, we can choose to 
build it only within the alpha-hull of the set of towns.

```{r}
dat_roll <- rollply(dat, ~ x + y, wdw.size = 1e4,
                    grid_npts = 10e3, # number of grid points 
                    grid_type = "ahull_fill", # grid type: fills an alpha-hull with the given number of points
                    grid_opts = list(alpha = .05, # shape parameter of the hull
                                     verbose = TRUE), 
                    how_many_with_name, regexp = "^ker")
```

Note that building an alpha hull-based grid can be quite computationally 
expensive, as it requires iterating to find the suitable number of points. 
However, one can pregenerate grids using the `build_grid_*` functions family and 
supply them directly to rollply using the `grid` argument, so this computation 
can be only done once (see below). 

So, are there really more town named *ker-*something in Brittany than elsewhere?
I'll let you judge (spoiler: yes!):

```{r, fig.width=7, fig.height=6}
ggplot(dat_roll) +
  geom_raster(aes(x, y, fill = ker)) +
  scale_fill_distiller(palette = 'Greys')
```

# Building grids: helper functions

As seen in the french towns example, rollply uses internally a grid of
coordinates. For each points of this grid it selects the observations within the
window, then applies the function on this subset. The user can either provide a
grid as a `data.frame` or rollply will take care of building one automatically.

Several helper functions are provided to build nice grids, they all start
with *build_grid_*.

- *build_grid_identical* builds a grid with an identical number of points on 
each dimension

- *build_grid_squaretile* (2D only) builds a grid of points with square tiles 
(same distance in X and Y between each points)

- *build_grid_ahull_crop* (2D only) builds a grid of points with square tiles, 
then discard all the points that do not fall in the alpha-hull of the actual 
data.

- *build_grid_ahull_fill* (2D only) same as above, but iteratively tries to 
build a grid with a final number of points approximately equal to 
the requested number of points (parameter `grid_npts`).


```{r, echo=FALSE, fig.width=4*2.5, fig.height=4}
data(meadow)

# We project lon/lat to UTM zone 11 (southern california)
meadow[ ,c('x','y')] <- proj4::project(as.matrix(meadow[ ,c('lon','lat')]),
                                       "+proj=utm +zone=11 +ellps=WGS84")

ggplot(meadow, aes(x,y)) +
  geom_point(shape='+') +
  ggtitle('Sample points') +
  xlab('UTM X') +
  ylab('UTM Y')
```

For this example, we will use samples from a vegetation survey in a meadow
in Yosemite National Park, California.

```{r}
# We request a grid with approximately this number of points:
npts <- 500
base.plot <- ggplot(NULL, aes(x,y)) +
               geom_point(data = meadow, shape='+') +
               xlab('UTM X') +
               ylab('UTM Y') 

grids <- list(identical  = build_grid_identical(meadow[ ,c('x','y')], npts),
              squaretile = build_grid_squaretile(meadow[ ,c('x','y')], npts),
              ahull_crop = build_grid_ahull_crop(meadow[ ,c('x','y')], npts),
              ahull_fill = build_grid_ahull_fill(meadow[ ,c('x','y')], npts))

plot_grid <- function(grid_type) {
  base.plot +
    geom_point(data=grids[[grid_type]]) +
    annotate('text', x = min(meadow$x), y = min(meadow$y),
            label = paste(nrow(grids[[grid_type]]), "points"),
            hjust = 0, vjust = 1)
}

```

- *build_grid_identical*:

```{r, fig.width=4*2.5, fig.height=4}
plot_grid('identical')
```

- *build_grid_squaretile* (mind the change of the figure aspect ratio: tiles are
squares):

```{r, fig.width=4*2.5, fig.height=4*2.5/4.92}
plot_grid('squaretile')
```

- *build_grid_ahull_crop*:

```{r, fig.width=4*2.5, fig.height=4*2.5/4.92}
plot_grid('ahull_crop')
```

- *build_grid_ahull_fill*: 

```{r, fig.width=4*2.5, fig.height=4*2.5/4.92}
plot_grid('ahull_fill')
```

Note that the outline as given by the alpha-hull of a set of points depends on 
a parameter, `alpha` that determines its intrincateness. A reasonable default
is provided, but it is a good idea to try different values to get an idea of 
the best fit. For more information on the alpha hull, please refer to the 
[package](https://cran.r-project.org/web/packages/alphahull/alphahull.pdf) 
documentation. 

# Notes

## Performance considerations

`rollply` inherits plyr's pros and cons. As in the latter's functions, 
parallelism or progress report is just one argument away (set `.parallel = TRUE` 
or `.progress = "time"`). However, rollply does a lot of `data.frame` subsetting 
which remains an expensive operation in R. 

## Bugs & comments

rollply has well-known bugs and is still in active development! Development 
happens at [github](https://github.com/alexgenin/rollply). Do not hesitate to 
post issues or pull requests.
