---
title: "A visual introduction to rollply"
author: "Alexandre Génin <alex@lecairn.org>"
output:
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{rollply}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# A visual introduction to rollply


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r, output = "none", messages = FALSE}
library(proj4) 
library(ggplot2) 
library(rgdal)
library(tidyr) 
library(rollply) 
library(plyr) 
```

Rollply is a small function built upon [plyr](http://plyr.had.co.nz/)'s `ddply` 
function to facilitate moving-window-based computations. If you have a 
`data.frame`, give the dimensions over which the function the window should 
move, and rollply will make the subsets, apply the function on them and then 
combine the results.

In short, rollply extends the 
[split-apply-combine strategy](http://plyr.had.co.nz/) to moving-window 
computations, using a similar syntax. This tutorial thus assumes some basic 
familiarity with `plyr`. Let's start with a simple example.

# Examples

## A time-series example

A simple use of moving-windows is adding a trendline to a time series plot.
We will use the CO2 data from the Mauna Loa [NOAA
Observatory](http://www.esrl.noaa.gov/gmd/ccgg/trends/) as a
environmentally-conscious example.

```{r, fig.width=7, fig.height=4}
# Download and format data
url <- "ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt"
hawaii <- read.table(url)[ ,c(3,4)]
names(hawaii) <- c('date','CO2')
hawaii[hawaii$CO2 < 0, "CO2"] <- NA # mark NAs as such

# Display original trend
CO2.plot <- ggplot(hawaii) + geom_line(aes(date, CO2)) + ylab("CO2 (ppm)")
print(CO2.plot)
```

There is a clear trend here! Let's smooth out the season effect (the wiggles
in the black curve). We'll use a window with a size of one year to compute a
yearly average.

```{r, fig.width=7, fig.height=4}
# with smoothed trend
hawaii.smoothed <- rollply(hawaii, ~ date, wdw.size = 1,
                           summarize, CO2.mean = mean(CO2, na.rm = TRUE), )
CO2.plot + geom_line(aes(date, CO2.mean), data = hawaii.smoothed, color = 'red')
```

And voilà! A rather nice, although a bit depressing, trend line for our data. 
When working on time-series, this represents an alternative to specialized 
packages such as `zoo` that also provide tools to apply functions in rolling
windows. 

Let's take a more complex example that works on two-dimensional data.

## Exploring french town names

If you open a map of France, you'll notice that towns and villages
tend to have names that follow patterns. For instance, Brittany's towns are
famous for having names starting with a *"ker-"*. Many towns in
[Lorraine](https://en.wikipedia.org/wiki/Lorraine_%28region%29) end in *"-ange"*
(a legacy from the german ending *"-ingen"*).

Can we visually explore the distribution of town names ?

A moving-window approach essentially boils down to the following steps:

1. Build a grid over the whole country
2. For each point of the grid, take all the towns that are less than xx
kilometers from it.
3. Check the names of the towns and count the ones matching a pattern. Return
the number of matching towns.
4. Combine the results in a `data.frame`

Like `ddply` (in package `plyr`), `rollply` takes care of points 1,2 and 4. We
just need to define a function that does number 3!
 
Let's download a [dataset](http://www.nosdonnees.fr/dataset/listes-des-communes-par-r-gions-d-partements-circonscriptions) of town names with their geographical coordinates:

```{r, warning=FALSE, results="hide"}
# Download and prepare dataset
# Source and decription: 
# https://publicdata.eu/dataset/listes-des-communes-par-rgions-dpartements-circonscriptions

tmpfile <- tempfile()
url <- paste0('http://www.nosdonnees.fr/wiki/images/b/b5/',
         'EUCircos_Regions_departements_circonscriptions_communes_gps.csv.gz')
download.file(url, destfile = tmpfile)
dat <- read.csv2(tmpfile, stringsAsFactors = FALSE)
file.remove(tmpfile)
dat <- dat[with(dat, latitude != "" | ! grepl(",", latitude)), 
           c('nom_commune', 'latitude', 'longitude')]
colnames(dat) <- c('name', 'latitude', 'longitude')

dat[ ,'name']      <- as.factor(tolower(dat[ ,'name']))
dat[ ,'latitude']  <- as.numeric(dat[ ,'latitude'])
dat[ ,'longitude'] <- as.numeric(dat[ ,'longitude'])

# We use an equirectangular projection to work on true distances
dat <- na.omit(dat)
dat <- data.frame(dat, proj4::project(dat[ ,c('longitude','latitude')],
                                      '+proj=eqc'))
dat <- dat[ ,c('name','x','y')]

```

```{r, fig.width=7, fig.height=6}
# Visualise distribution of towns
str(dat)
ggplot(dat) + geom_point(aes(x, y), alpha=.1)
```

Nice, let's see whether *"ker"*-named towns are predominantly in Brittany.

```{r, fig.width=7, fig.height=6}
# This is our custom function.
how_many_with_name <- function(regxp, X) sum(grepl(regxp, X))


dat_roll <- rollply(dat, ~ x + y, wdw.size = 1e4, grid_npts = 10e3,
                    summarize, ker = how_many_with_name("^ker", name))

ggplot(dat_roll) +
  geom_raster(aes(x, y, fill = ker)) +
  scale_fill_distiller(palette = 'Greys')
```

It seems there are indeed many towns with a name starting with "*ker*" in
Brittany (and a couple in Alsace/Lorraine, too). However, our plot is pretty
ugly: we can't see the actual country shape! When nothing is specified,
rollply computes its values over a rectangular grid than spans the maximum width
and height of the original dataset.

Here, our spatial distribution of towns reflect pretty well the overall shape of 
the country, so instead of a rectangular grid we can build a grid inside the 
overall outline (the alpha-hull) of the dataset. Another option would have been 
to provide a polygon of the country to clip the grid of points.

```{r}
dat_roll <- rollply(dat, ~ x + y, wdw.size = 1e4,
                    grid_npts = 10e3,
                    grid_type = "ahull_fill",
                    grid_opts = list(alpha = .05, # shape parameter for hull
                                     verbose = TRUE),
                    summarize, ker = how_many_with_name("^ker", name))
```

Note that building an alpha hull-based grid can be quite slow, as it is built on 
the pure-R `alphahull` package. However, one can pregenerate grids using the 
`build_grid_*` functions family and supply them directly to rollply using the 
`grid` argument, so this computation can be only done once (see below).

So, are there really more town named *ker-*something in Brittany than elsewhere?
I'll let you judge (spoiler: yes!):

```{r, fig.width=7, fig.height=6}
ggplot(dat_roll) +
  geom_raster(aes(x, y, fill = ker)) +
  scale_fill_distiller(palette = 'Greys')
```

# Building grids: helper functions

As seen in the french towns example, rollply uses internally a grid of
coordinates. For each points of this grid it selects the observations within the
window, then applies the function on this subset. The user can either provide a
grid as a `data.frame` or rollply will take care of building one automatically.

Several helper functions are provided to build nice grids, they all start
with *build_grid_*.

- *build_grid_identical* builds a grid with as many points on each dimension

- *build_grid_squaretile* (2D only) builds a grid of points with square tiles

- *build_grid_ahull_crop* (2D only) builds a grid of points with square tiles, 
then discard all the points that do not fall in the alpha-hull of the actual 
data.

- *build_grid_ahull_fill* (2D only) same as above, but iteratively tries to 
build a grid with a final number of points approximately equal to the parameter `grid_npts`.


```{r, echo=FALSE, fig.width=4*2.5, fig.height=4}
data(meadow)

# We project lon/lat to UTM zone 11 (southern california)
meadow[ ,c('x','y')] <- proj4::project(as.matrix(meadow[ ,c('lon','lat')]),
                                       "+proj=utm +zone=11 +ellps=WGS84")

ggplot(meadow, aes(x,y)) +
  geom_point(shape='+') +
  ggtitle('Sample points') +
  xlab('UTM X') +
  ylab('UTM Y')
```

For this example, we will use samples from a vegetation survey in a meadow
in Yosemite National Park, California.

```{r}
# We request a grid with approximately this number of points:
npts <- 500
base.plot <- ggplot(NULL, aes(x,y)) +
               geom_point(data = meadow, shape='+') +
               xlab('UTM X') +
               ylab('UTM Y')

grids <- list(identical  = build_grid_identical(meadow[ ,c('x','y')], npts),
              squaretile = build_grid_squaretile(meadow[ ,c('x','y')], npts),
              ahull_crop = build_grid_ahull_crop(meadow[ ,c('x','y')], npts),
              ahull_fill = build_grid_ahull_fill(meadow[ ,c('x','y')], npts))

plot_grid <- function(grid_type) {
  base.plot +
    geom_point(data=grids[[grid_type]]) +
    annotate('text', x = min(meadow$x), y = min(meadow$y),
            label = paste(nrow(grids[[grid_type]]), "points"),
            hjust = 0, vjust = 1)
}

```

- *build_grid_identical*:

```{r, fig.width=4*2.5, fig.height=4}
plot_grid('identical')
```

- *build_grid_squaretile* (mind the change of the figure aspect ratio: tiles are
squares):

```{r, fig.width=4*2.5, fig.height=4*2.5/4.92}
plot_grid('squaretile')
```

- *build_grid_ahull_crop*:

```{r, fig.width=4*2.5, fig.height=4}
plot_grid('ahull_crop')
```

- *build_grid_ahull_fill*: 

```{r, fig.width=4*2.5, fig.height=4}
plot_grid('ahull_fill')
```

Note that the outline as given by the alpha-hull of a set of points depends on 
a parameter, `alpha` that determines its intrincateness. A reasonable default
is provided, but it is a good idea to try different values to get an idea of 
the best fit. For more information on the alpha hull, please refer to the 
[package](https://cran.r-project.org/web/packages/alphahull/alphahull.pdf) 
documentation. 

# Notes

## Performance considerations

`rollply` inherits plyr's pros and cons. As in the latter's functions, 
parallelism or progress report is just one argument away (set `.parallel = TRUE` 
or `.progress = "time"`). However, rollply does a lot of `data.frame` subsetting 
which remains an expensive operation in R. 

## Bugs & comments

rollply is my first package, has well-known bugs and is still in active
development!

Development happens at [github](https://github.com/alexgenin/rollply). Do not
hesitate to post issues or pull requests.
