---
title: "Building grids within rollply"
author: "Alexandre Génin"
date: "08/06/2015"
---

```{r, echo=FALSE, fig.width=2*4.924665, fig.height=4}
library(ggplot2)
library(rgdal)
library(rollply)

load('./meadow_example.dat')
mdw <- yosemite_meadow[ , c('lon','lat','meanvwc','tcover')]

# We project lon/lat to UTM zone 11 (southern california)
mdw[ ,c('x','y')] <- project(as.matrix(mdw[ ,c('lon','lat')]), 
                             "+proj=utm +zone=11 +ellps=WGS84")  

ggplot(mdw, aes(x,y)) + 
  geom_point(shape='+') + 
  ggtitle('Point surveyed in a meadow') + 
  xlab('UTM X') + 
  ylab('UTM Y')

```

Rollply uses internally a grid of coordinates. For each points of this
grid it selects the observations falling within the window, then applies the function using this subset. The user can either provide a grid as a data.frame or rollply will take care of building one automatically. 

Several helper functions are provided to build nice 2D grids, they all start with *build_grid_*.

- *build_grid_identical*: builds a grid of regularly-spaced points, with the same number of points on each dimension.

- *build_grid_proportional* (2D only): build a 2D grid of points, with a number of points on each dimension proportional to the total span of that dimension in the 
observations.

- *build_grid_ahull_crop* (2D only): builds a 2D grid of points, then discard   all the points that do not fall in the alphahull of the actual data available.

- *build_grid_ahull_fill* (2D only): same as above, except iterating to build a grid with a final number of points approximately equal to the one asked for (parameter grid.res in function)

<!-- to link: http://CRAN.R-project.org/package=alphahull -->

In images, let's say we want a grid of 500 points to compute something on the 
meadow presented above.

```{r} 
# We request a grid with approximately this number of points
npts <- 500
base.plot <- ggplot(NULL, aes(x,y)) + 
               geom_point(data=mdw, shape='+') +
               xlab('UTM X') + 
               ylab('UTM Y')

grids <- list(grid_identical     = build_grid_identical(mdw[ ,c('x','y')], npts),
               grid_proportional = build_grid_proportional(mdw[ ,c('x','y')], npts),
               ahull_crop        = build_grid_ahull_crop(mdw[ ,c('x','y')], npts),
               ahull_fill        = build_grid_ahull_fill(mdw[ ,c('x','y')], npts))
```

- *build_grid_identical*:

```{r, fig.width=2*4.924665, fig.height=4}
nrow(grids[['grid_identical']]) # Total number of points in grid
base.plot + geom_point(data=grids[['grid_identical']])
```

- *build_grid_proportional*:

```{r, fig.width=2*4.924665, fig.height=4}
nrow(grids[['grid_proportional']]) # Total number of points in grid
base.plot + geom_point(data=grids[['grid_proportional']])
```

- *build_grid_ahull_crop*:

```{r, fig.width=2*4.924665, fig.height=4}
nrow(grids[['ahull_crop']]) # Total number of points in grid
base.plot + geom_point(data=grids[['ahull_crop']])
```

- *build_grid_ahull_fill*: takes much longer to run but gives a number of points that correspond approximately to what was requested

```{r, fig.width=2*4.924665, fig.height=4}
nrow(grids[['ahull_fill']]) # Total number of points in grid
base.plot + geom_point(data=grids[['ahull_fill']])
```
